Scaling up a GF grammar and preparing it for cool stuff for Swedish

### 2011-11-30 doc/report/report.tex
report

\subsubsection*{Reflexive pronouns}
The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog \textbf{sig}. & Han såg \textbf{sitt} barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog \textbf{honom}. & Han såg \textbf{hans} barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

@@

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about modern
Swedish and the words are equipped with both sematic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of sematically connected words.\\
Example of katt? \\
For our purpose however, only the morphological data was needed.\\
Example of morpho for katt.\\
Saldo provides this in XML format under LPGL. The data can be processed and
translated to GF format as described in 
section \ref{sec:prog.saldo}.
%Other usages?

\subsection{GF}
The main component of the the project is the Grammatical Framework\ref{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming langauges
like Haskell\footnote{}, ML\footnote{}, Prolog\footnote{} and Agda\footnote{}. Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

Being developed with multilinguality in mind, GF makes an important distiction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages, it models the semantics of the domain without blanda in
language specific features such as agreement or word order.

@@

  fun 
    Pred : N -> V -> S ;
    cat_N : N ;
    sleep_V : V ;
 }
\end{verbatim}
\caption{A small example of an abstract syntax}
\end{figure}

The example shows an abstract grammar defining three categories, % \verb|Categories|,
one for nouns, one for verbs and one for sentences. An abstract grammar also gives
the types of functions. In this case we have \verb|Pred| which
tells us that by taking a noun and a verb we can form a sentence. At this stage, 
no information of how this is done is given. The grammars also defines two words,
\verb|cat| and \verb|sleep|.

@@

Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:
\verb|Pred cat sleep = "katten sover"|.\\
Let's look at a more complicated example, where nouns can be used in both
plural and singular. To do this, we add a category \verb|NP| to the
abstract. \verb|N| now means a noun
in any of the number, and we introduce two funciton, \verb|NSg : N -> NP| and
\verb|NPl : N -> Pl| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

If we want to implement a English version of the grammar, we encounter another
problem: the verb form depends on the noun phrase. The \verb|NP| needs to carry
information about which number it is in, and \verb|Pred| can pass this on to the
verb. Hence, the \verb|V| also needs to be a table, depending on the number.
\begin{figure}[h!]
\begin{verbatim}              
concrete TestGrammarEng of TestGrammar = {
  lincat S = Str ;
         V = {s : Num => Str} ;
         N = {s : Num => Str} ;
         NP = {s : Str ; num = Num} ;
  lin   
    Pred n v = n.s ++ v.s ! n.num ;
    NPl n = {n.s ! Pl ; num = Pl} ;
    NSg n = {n.s ! Sg ; num = Sp} ;
    cat_N = {s = table {Sg => "the cat" ;
                        Pl => "the cats"}};
    sleep_V = {s = table {Sg => "sleeps" ; 
                          Pl => "sleep"}};
  param Num = Sg | Pl ;
  }
\end{verbatim}           
\caption{English implementation}
\label{fig:gfTestEng}
\end{figure}

@@

We now have two implementations of the abstract. The resulting GF grammar is able to
both parse a string to the abstract tree and to go in the other direction; to produce
a string of natural language given a abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

Valency information is normally built into the GF categories. Add V2 to the example.
Point out how N + V can be glued together to a string, the components cannot be seen
anymore.

@@

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} %correct!!!
\label{gfSwe:parseable}
The grammar can construct and parse the sentence in example \ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples.pdf}
\caption{Parse tree for ``Han har inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

@@

%% Clauses
\subsubsection*{Presentering, Impersonal constructions, Topicalisering}
Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{
*Det äter en fågel på taket\\
 ''There is a bird eating on the roof"}

@@

There are also restrictions on the subject.
\enumsentence{\begin{tabular}{ll}
*Det sitter den fågeln på taket. & *Det sitter fågeln på taket.\\
``There are that bird sitting on the roof'' & ``There are the bird sitting on the roof'' \\
\end{tabular}}
The noun must be in indefinite form, even though some determiners are allowed.
\enumsentence{Det sitter några fåglar på taket\\
              ``There are some birds on the roof''}
\enumsentence{*Det sitter fåglarna på taket\\
``There are the birds on the roof''}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The GF grammar does not make any distinction between indefinite and definite noun
phrases. Therefore this is handled by looking at the determiner. \\
\verb|FormalSub : SimpleVP -> Det -> CN -> Cl ;|\\
The function combines a verb phrase a determiner and a noun phrase in to a clause.
If the determiner requires a noun in determinened form, the clause is put to NONEXIST.
However, this is not an ideal solution. There are other examples of when we
need to know the definitness of the noun phrase, for example when using
the reflexive possessive pronoun `sin' (see section `Reflexive pronouns' below)
or when using some predeterminers (see section `Quantifiers').

@@

\subsubsection*{Reflexive pronouns}
%varandra and sin/sitt/sig.
Both sig and sin
Not subject, but NP.
Want the following:
\enumsentence{Han såg sina barns skor.}
\enumsentence{Sina vantar hade han glömt på tåget.}
but not
\enumsentence{*Sina vantar var kvar på tåget.}
\enumsentence{Han ger sina pengar till sina syskon.}
\enumsentence{*Jag ger sina pengar till sina syskon.}
\enumsentence{Han är längre än sin kompis.}
\enumsentence{Hon tyckte om alla sina elever.}
\enumsentence{Han såg sin hund och sitt hus}
\label{ex:refGen}
\enumsentence{Han är längre än sin kompis}
\label{ex:ReflAP}
% hon ber sin katt gå - sin subject in clause.

Suggest a connection between the subject and the object.
Could be combined with 'sig', a ReflNP without another noun.
-- man ser sig!!
Make a distinction between objects and subjects?
do not want to copy all rules for another category. Would like depent types!
work pretty well by letting all NP depend on the subject in the phrase.
The current version is nice and works for recursive genitivs!
Extra info, but maybe this is just the case about NPs.
Discuss how other solutions, NPR and Predet : Predet -> CN -> NPR ;
*Slash* : VPSlash -> NPR -> VP ;
Keep number of fields down, but introduces another category, and one more
label to the parse tree, is it interesting? How to combine it with genitive 
without making it ambiguous (Han såg barnens hund).
What if we add alot of other rules that uses NP, will they all need to be
duplicated then?

'han såg sin mammas bil', they same could be applied to
APs for example \ref{ex:ReflAP} or Adv for
\enumsentence{Han leker oftare än sin kompis}
 (ComparAdvAdj)
Atm you can't focus this kind of object, then a ClSlash need info about the
subject. FunRP also tricky (and what does it mean, we can say
Han såg sin katt i vilken musen låg), in this case RP needs to give more information.
Kort sagt, the information about the subject must be spread
in different parts of the sentence. 
Could also decide to always produc 'sin' -> jag såg sin bil
to avoid the dependence, but not nice. Or keep a flag in NP saying if it is
in third person, and add nonexist for others. But this way we can also get rid
of ReflVP and the possessive prons \verb|PossPron| should only be used for somebody else:
'jag såg din bil'.
Unfortunaly, have to change ReflVP, this only allows this in object for VPSlash, not FocObj,
not 'alla sina', usw. Want it to handle conjuction as in example \ref{ex:refGen}
etc.
varandra, works the same but has no case for Sg.
Johan vill att jag rakar sig* -> honom  
%%

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)
\subsubsection*{Relative clauses}
The resource grammar already provided a number of ways to express relative and embedded clauses, such as
\enumsentence{Frågan är vilken färg den har\\
                   Pojken, som är blyg, tystnar\\
                   Han såg kunden som tyckte om sallad\\
                   Jag tänkte på huset i vilket hon bodde} % better example!!

>>>>

\subsubsection*{Reflexive pronouns}
\label{swe:refl}
The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog \textbf{sig}. & Han såg \textbf{sitt} barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog \textbf{honom}. & Han såg \textbf{hans} barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

@@

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with both sematic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of sematically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}
For our purpose however, only the morphological data was needed.
Saldo provides this in XML format under LPGL. The data can be processed and
translated to GF format as described in 
section \ref{sec:prog.saldo}.
%Other usages?

\subsection{Grammatical Framework}
The main component of the the project is the Grammatical Framework\ref{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming langauges
like Haskell\footnote{}, ML\footnote{}, Prolog\footnote{} and Agda\footnote{}. Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

An important feature of GF is
%Being developed with multilinguality in mind, GF makes an important 
the distiction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages, it models the semantics of the domain without concern for
language specific features such as agreement or word order.

@@

  fun 
    Pred : N -> V -> S ;
    cat_N : N ;
    sleep_V : V ;
 }
\end{verbatim}
\caption{A small example of an abstract syntax}
\label{fig:gfAbstract1}
\end{figure}

The example in figure \ref{fig:gfAbstract1} shows an abstract grammar defining three categories, % \verb|Categories|,
one for nouns, one for verbs and one for sentences. An abstract grammar also gives
the types of functions. In our case we have \verb|Pred| which
tells us that by taking a noun and a verb we can form a sentence. At this stage, 
no information of how this is done is given. The grammars also defines two words,
the noun \verb|cat| and the verb \verb|sleep|.

@@

Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. To do this, we add a category \verb|NP| to the
abstract. \verb|N| now means a noun
in any of the number, and we introduce two funciton, \verb|NSg : N -> NP| and
\verb|NPl : N -> Pl| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

If we want to implement a English version of the grammar, we encounter another
problem: the verb form depends on the noun phrase. The \verb|NP| needs to carry
information about which number it is in, and \verb|Pred| can pass this on to the
verb. Hence, the \verb|V| also needs to be a table, depending on the number.
\begin{figure}[h!]
\begin{verbatim}              
concrete TestGrammarEng of TestGrammar = {
  lincat S = Str ;
         V = {s : Num => Str} ;
         N = {s : Num => Str} ;
         NP = {s : Str ; num : Num} ;
  lin   
    Pred n v = n.s ++ v.s ! n.num ;
    NPl n = {s = n.s ! Pl ; num = Pl} ;
    NSg n = {s = n.s ! Sg ; num = Sg} ;
    cat_N = {s = table {Sg => "the cat" ;
                        Pl => "the cats"}};
    sleep_V = {s = table {Sg => "sleeps" ; 
                          Pl => "sleep"}};
  param Num = Sg | Pl ;
  }
\end{verbatim}           
\caption{English implementation}
\label{fig:gfTestEng}
\end{figure}

@@

We now have two implementations of the abstract. The resulting GF grammar is able to
both parse a string to a abstract tree and to go in the other direction; to produce
a string of natural language given an abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

Valency information is normally built into the GF categories. To indicate that
a verb is transitive and needs an object, we use the category \verb_V2_.
\begin{verbatim}
abstract:
cat V2 ; 
eat_V2 : V2 ;
rat_N ;
UseV2 : V2 -> NP -> V ;

concrete:
eat_V2 = "äta" ;
rat_N = {s = table {Sg => "råttan" ;
                    Pl => "råttorna"}};
UseV2 v np = v ++ np ;
\end{verbatim}

The example above now allows ``Katten äter råttan", but not ``Katten äter".
\verb|UseV2| concatenate the strings ``äter" and ``råttan". Note that if we
wanted to add negation, ``katten äter inte råttan", or question "äter katten råttan",
we cannot glue the verb to the object at once. Instead we would have to use a table,
with one field for the verb and one for the object.

@@

\subsection{The Swedish resource grammar}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} %correct!!!
\label{gfSwe:parseable}
The grammar can construct and parse the sentence in example \ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples.pdf}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

@@

%% Clauses
\subsubsection{Impersonal constructions}%Presentering, , Topicalisering
\label{sec:Formal}
Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{
*Det äter en fågel på taket\\
 ''There is a bird eating on the roof"}

@@

There are also restrictions on the subject.
\enumsentence{\begin{tabular}{ll}
*Det sitter den fågeln på taket. & *Det sitter fågeln på taket.\\
``There are that bird sitting on the roof'' & ``There are the bird sitting on the roof'' \\
\end{tabular}}
The noun must be in indefinite form, even though some determiners are allowed.
\enumsentence{Det sitter några fåglar på taket\\
              ``There are some birds on the roof''}
\enumsentence{*Det sitter fåglarna på taket\\
``There are the birds on the roof''}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The GF grammar does not make any distinction between indefinite and definite noun
phrases. Therefore this is handled by looking at the determiner. \\
\verb|FormalSub : SimpleVP -> Det -> CN -> Cl ;|\\
The function combines a verb phrase a determiner and a noun phrase in to a clause.
If the determiner requires a noun in determinened form, the clause is put to NONEXIST.
However, this is not an ideal solution. There are other examples of when we
need to know the definitness of the noun phrase, for example when using
the reflexive possessive pronoun `sin' (see section `Reflexive pronouns' below)
or when using some predeterminers (see section `Quantifiers'). Not true, not the same :(

@@

\subsubsection{Reflexive pronouns}
%varandra and sin/sitt/sig.
An important piece in a Swedish grammar is the reflexive pronouns and
%both \emph{sin} and , eg. \emph{sin}.
the possessive pronouns as described in section \ref{swe:refl}.
Those make up noun phrases wich are not allowed in subjects, 
and we want to be able to parse to following sentences:
\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw his children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
His gloves, he had forgotten on the train.}
\enumsentence{Han ger sina pengar till sina syskon.\\
He gives his money to his children.}
\enumsentence{Han är längre än sin kompis.\\
He is taller than his friend.} \label{ex:ReflAP}
\vspace{-2mm}
\enumsentence{Han är här oftare än sin kompis.\\
He is here more often than his friend.} \label{ex:ReflAdv}
\vspace{-2mm}
\enumsentence{Hon tyckte om skolan och alla sina elever.\\
She liked the school and all her students.}\label{ex:ReflConj}
\enumsentence{Han såg sina få böcker och sin penna.\\
He saw his few books and his pencil.}\label{ex:ReflConj2}
\vspace{-2mm}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all her friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave.}
but not
\enumsentence{*Sina vantar var kvar på tåget.\\
Her gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give his money to his siblings. }
\enumsentence{*Han vill att jag rakar sig.\\
He wants me to shave himself.}

The reflexive pronouns require another subject whith which they agree in
number and person. Apart from this restriction, they may be used as normal
noun phrases and conjuncted (sentence \ref{ex:ReflConj} and \ref{ex:ReflConj2}),
determined (sentence \ref{ex:ReflPredet}) and used as genitives (sentence
\ref{ex:ReflGen}) as such.
%Could be combined with 'sig', a ReflNP without another noun.
In short, what is wanted can be summerized as follows:
\begin{figure}
\begin{enumerate}
\item{
Construct a noun phrase from a common noun phrase \\
katt $\rightarrow$ sin katt} % \\
\item{Modify it like a \verb|NP|\\
sina katter $\rightarrow$ alla sina katter}
\item{Use it only as object\\
sin katt $\rightarrow$ han såg sin katt }
\end{enumerate}
\caption{Rules for noun phrases with reflexive possessive pronouns}
\label{fig:reflrestrict}
\end{figure}
If we ignore the second requirement, special functions for using common nouns
together with reflexive possessive pronouns could be introduced:
\begin{verbatim}
ReflVP: CN -> VPSlash -> VP ;
        mat + äta     -> äta sin mat 
\end{verbatim}
This way, using the phrases as subject is excluded, but neither of sentence
\ref{ex:ReflConj}, \ref{ex:ReflConj2} or \ref{ex:ReflPredet} is allowed.\\
The simplest way of fixing this is to treat the phrases as normal \verb_NP_s.
We can add a rule for using the possessives with common noun phrases:
\begin{verbatim}
ReflNP: CN -> NP ;
        mat -> sin mat 
PredetNP : Predet -> NP       -> NP ;
           alla   + sina barn -> alla sina barn
\end{verbatim}
But by keeping them as \verb_NP_s we lose information, and there is no way of
keeping them from being used as subjects. We have ignored the third restriction
and are allowing sentences like \ref{ex:ReflSubj}.

The solution introduces a new object type, \verb_Obj_, which is
identical to normal \verb_NP_ except for that it depends on the agreement of
the subject. All functions using noun phrase as objects are changed to dealing
with objects instead.
\begin{verbatim}
ReflVP: Obj     -> VPSlash -> VP ;
     -- sin mat +  äta     -> äta sin mat 
NPtoObj : NP  -> Obj ;
PredVP  : NP -> VP -> Cl ;  -- NP used as subject
\end{verbatim}
Any normal noun phrase may also be used as an object, whereas subjects only
can be made up by \verb-NP-s.
Since the resource grammar did not make any distinction between
subjects and objects, the new Swedish grammar needed to be separated from the old
structure. 
Further, the dependence on the subject needs to be applied to adverbial phrases and
adjective phrases as well. The adjetive phrase in sentence \ref{ex:ReflAP} and
the adverbial phrase in sentence \ref{ex:ReflAdv} are examples of this; if the
subject was changed to 2nd person, they  would need to change too.
\enumsentence{\textbf{Du} är längre än \textbf{din} kompis.\\
You are taller than your friend.} \label{ex:ReflAP2}
\vspace{-2mm}
\enumsentence{*Du är här oftare än sin kompis.\\
You are here more often than his friend.} \label{ex:ReflAdv2}
Adverbial phrases may also be parts of new noun phrases
\enumsentence{\shortex{3}
{Taket & på sitt hus &}
{\scriptsize{\textbf{NP +}} & \scriptsize{\textbf{(Adv with Obj)}} & $\Rightarrow$ \scriptsize{\textbf{Obj}} }
{The roof of his house}}
\enumsentence{\shortex{3}
{Sitt hus & på berget &}
{\scriptsize{\textbf{Obj +}} & \scriptsize{\textbf{Adv}} & $\Rightarrow$ \scriptsize{\textbf{Obj}} }
{His house on the mountain}}

The first restriction in figure \ref{fig:reflrestrict} should be slightly
modified, to allow certain kind of \verb|NP|s to be used.
\begin{figure}[h]
\begin{enumerate}
\item{
Construct a noun phrase from a common noun phrase or certain noun phrases \\
få katter $\rightarrow$ sina få katter} 
\end{enumerate}
\caption{Modification of \ref{fig:reflrestrict}}
\end{figure}

The kind of noun phrases that can be used seems to coincide with the ones
that can be used with formal subjects (see section `Impersonal constructions').

% agreement to subj: Also good for double functions : Flickan, hon var inte dum, hon?
% F1,2,3 : NP -> VP -> Cl ; NP needs gender (Masc,Fem,Utr,Neutr)

%Make a distinction between objects and subjects?
%do not want to copy all rules for another category. Would like depent types!
%work pretty well by letting all NP depend on the subject in the phrase.
%The current version is nice and works for recursive genitivs!
%Extra info, but maybe this is just the case about NPs.
%Discuss how other solutions, NPR and Predet : Predet -> CN -> NPR ;
%*Slash* : VPSlash -> NPR -> VP ;
%Keep number of fields down, but introduces another category, and one more
%label to the parse tree, is it interesting? How to combine it with genitive 
%without making it ambiguous (Han såg barnens hund).
%What if we add alot of other rules that uses NP, will they all need to be
%duplicated then?

%'han såg sin mammas bil', they same could be applied to
%APs for example \ref{ex:ReflAP} or Adv for
%\enumsentence{Han leker oftare än sin kompis}
% (ComparAdvAdj)
%Atm you can't focus this kind of object, then a ClSlash need info about the
%subject. FunRP also tricky (and what does it mean, we can say
%Han såg sin katt i vilken musen låg), in this case RP needs to give more information.
%Kort sagt, the information about the subject must be spread
%in different parts of the sentence. 
%Could also decide to always produc 'sin' -> jag såg sin bil
%to avoid the dependence, but not nice. Or keep a flag in NP saying if it is
%in third person, and add nonexist for others. But this way we can also get rid
%of ReflVP and the possessive prons \verb|PossPron| should only be used for somebody else:
%'jag såg din bil'.
%Unfortunaly, have to change ReflVP, this only allows this in object for VPSlash, not FocObj,
%not 'alla sina', usw. Want it to handle conjuction as in example \ref{ex:refGen}
%etc.
%varandra, works the same but has no case for Sg.
%%

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)
\subsubsection{Relative clauses}
The resource grammar already provided a number of ways to express relative and embedded clauses, such as
\enumsentence{Frågan är vilken färg den har\\
                   Pojken, som är blyg, tystnar\\
                   Han såg kunden som tyckte om sallad\\
                   Jag tänkte på huset i vilket hon bodde} % better example!!
